A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-21 paper.tex
Cleanup
Update

% can see from the example (probably).

@@

% % If we want to use the coalescent, then really the assumptions
% % need to make sense.
% % n << Ne is a silly assumption in today's datasets
% A core assumption of the coalescent is that the sample size $n$
% is much less than the effective population size, $N_e$.
% Several human datasets now consist of hundreds of thousands of
% genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
% and so sample size is substantially \emph{larger} than $N_e$
% (often assumed to be $10^4$ in humans).
% Agricultural datasets are even more extreme.
% For example,
% the US dairy cattle database alone currently comprises more than 6 million
% animals with SNP array
% genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
% while the effective population size in modern dairy cattle breeds is
% less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
% An extreme sign of these breeding practices is that there are only two ancestral
% Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
% Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
% comprises close to 7000 genomes, which are part of multi-generation pedigrees
% with millions of animals and extensive SNP array genotype and phenotype
% data \citep[e.g.][]{Cesarani2022}.
% In another example, genome and SNP array data for 440,610 individuals within
% 7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
% were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
% Effective population size in modern pig breeds is also less than 100 due to
% intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
% Another core assumption of the coalescent model is that the genome (or
% at least the region under study) is short enough that the number of extant
% lineages remains much smaller than $N_e$ at all times. Whole
% genome sequences have been available for model organisms
% for over a decade now, % True? Citation?
% and indeed complete chromosome-level assemblies are possible
% in humans~\citep{miga2020telomere}.
% Projects are under way to obtain high-quality assemblies
% for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
% and ultimately worldwide~\citep{lewin2022earth}.

% While the coalescent as a model can be suprisingly robust to
% basic violations of the underlying assumptions~\citep{
% wakeley2012gene,bhaskar2014distortion,nelson2020accounting}
% it is reasonable to question the validity of the model in
% representing datasets such as UK Biobank.
% [Also mention somehow that $n\ll N_e$ is only one problem -
% we're also ignoring the myriad other evolutionary forces.]
% Such datasets cannot be reasonably be described by
% the coalescent as a statical model, and so more non-parametric
% methods like \tsinfer\ and \relate\ would seem more appropriate.
% [More]

% % Do downstream applications actually make use of such full ARGs?
% Downstream applications using estimated ARGs to date rarely
% make use of such complete recombination information.
% The \relate\ selection statistic~\citep{speidel2019method}
% computes a $p$-value for selection [FIXME] for each
% local tree independently by comparing the rates of coalescence
% at various time-points. For cases like strong positive selection,
% where we expect there to be a clear signal in the local tree,
% such tree-by-tree methods can be simple and powerful.
% In their method
% for estimating dispersal rates and the locations of genetic
% ancestors,
% \citep{osmond2021estimating} downsample trees along the genome
% so that they can be regarded as approximately independent.
% The SIA method for detecting selection~\citep{hejase2022deep}
% encodes local trees as a set of lineage counts at discrete
% time intervals, and uses these as feature for a
% type of machine learning algorithm
% that takes ``temporal'' correlations into account. Thus,
% while SIA does use some of the information about local tree correlation,
% clearly much of the detail about recombination
% events in a ``full'' ARG is lost.
% The \texttt{tsdate} algorithm (and related naive estimator
% for ancestral location) uses much more between-tree
% information~\citep{wohns2022unified}, explicitly using the gARG
% encoding to reason about ancestral nodes.
% A major strength of this approach is that it can be used on
% any gARG, ranging from estimates containing relatively little
% information about recombination (e.g., as estimated by \relate)
% to ``full'' ARGs containing complete recombination information
% (e.g., as estimated by \kwarg), and will fully utilise
% any between-tree correlation information that is present.

@@

@@

The amount of information about recombination that is inferrable
is the subject of much research
has a long history [summarise this in a couple of sentences?]
\cite{shipilina2023origin} discuss the idea of a ``haplotype block''
(equivalent to the ``bricked tree sequence'' of
\cite{nowbandegani2023extremely},
% TODO look this up. I expect it's basically the same??
and the block-based ARG data structure of
\citep{palamara2016argon}),
here we consider the unique sets of
samples that coalesce at a node in the ARG over a particular
genome interval, and the fundamental limitations they place
on ARG inference.

% [FIXME find somewhere natural for this. The last clause is too vague - what
% are the implications?]
% Alternative gARG encodings split transmitted intervals and associate each with
% a new node, resulting in a single genome being represented by a group of
% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
% efficiency of storage.

>>>>

@@

@@

%%%% Some old text about the coalescent we might still want to plunder
% % If we want to use the coalescent, then really the assumptions
% % need to make sense.
% % n << Ne is a silly assumption in today's datasets
% A core assumption of the coalescent is that the sample size $n$
% is much less than the effective population size, $N_e$.
% Several human datasets now consist of hundreds of thousands of
% genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
% and so sample size is substantially \emph{larger} than $N_e$
% (often assumed to be $10^4$ in humans).
% Agricultural datasets are even more extreme.
% For example,
% the US dairy cattle database alone currently comprises more than 6 million
% animals with SNP array
% genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
% while the effective population size in modern dairy cattle breeds is
% less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
% An extreme sign of these breeding practices is that there are only two ancestral
% Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
% Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
% comprises close to 7000 genomes, which are part of multi-generation pedigrees
% with millions of animals and extensive SNP array genotype and phenotype
% data \citep[e.g.][]{Cesarani2022}.
% In another example, genome and SNP array data for 440,610 individuals within
% 7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
% were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
% Effective population size in modern pig breeds is also less than 100 due to
% intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
% Another core assumption of the coalescent model is that the genome (or
% at least the region under study) is short enough that the number of extant
% lineages remains much smaller than $N_e$ at all times. Whole
% genome sequences have been available for model organisms
% for over a decade now, % True? Citation?
% and indeed complete chromosome-level assemblies are possible
% in humans~\citep{miga2020telomere}.
% Projects are under way to obtain high-quality assemblies
% for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
% and ultimately worldwide~\citep{lewin2022earth}.

@@

The amount of information about recombination that is inferrable
is the subject of much research
has a long history [summarise this in a couple of sentences?]
[Cite: \citep{myers2002detection,hayman2023recoverability}
% This might be where we mention Fisher's junctions and IBD?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/121
\cite{shipilina2023origin} discuss the idea of a ``haplotype block''
(equivalent to the ``bricked tree sequence'' of
\cite{nowbandegani2023extremely},
% TODO look this up. I expect it's basically the same??
and the block-based ARG data structure of
\citep{palamara2016argon}),
here we consider the unique sets of
samples that coalesce at a node in the ARG over a particular
genome interval, and the fundamental limitations they place
on ARG inference.

% [FIXME find somewhere natural for this. The last clause is too vague - what
% are the implications?]
% Alternative gARG encodings split transmitted intervals and associate each with
% a new node, resulting in a single genome being represented by a group of
% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
